Omics-based Single Step Trait Prediction

### 2017-03-27 paper/manuscript.tex
Init manuscript intro

% Introduction
\section{Introduction}

Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to hybrids, (2) explore the utility of single step 
predition with a quantitative predictor and (3) evaluate the impact of the
particular genetic space covered by the incomplete predictor on predictive
ability.

@@

\section*{Discussion}
\Genetics2level{Some discussion section}

@@

\subsection{More detailed discussion}

\Genetics2level{Another discussion section}

>>>>

% Introduction
\section{Introduction} 
% Necessity for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
multiple areas of genetic analyses such as animal and plant breeding
\cite{DeLosCampos2013} as well as disease diagnostics in humans 
\cite{DeLosCampos2010,Vazquez2016}.
The primary application of genomic prediction in breeding programs is for the 
accelerated selection of promising candidates without phenotypes.

% -   Describe the two step procedure and its shortcomings.
% Incomplete coverage of individuals with genomic data.
In animal breeding, only a small fraction of all animals that are recorded in
the pedigree also have genotypes.
To utilize all available phenotpyic information for model training methods were 
needed that would allow for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
In the early years of genomic prediction, a two-step procedure was the most
intuitive choice for establishing a new selection scheme.
The first step constists of conventionally estimating breeding values based on
pedigree relationships and therefore the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (\textit{i.e.}
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes. 
Those markers can then be used to predict genomic breeding values for solely
genotyped and potentially very young individuals.

% Shortcomings of the two-step procedure.
\begin{itemize}
  \item Bias introduced by genomic information giving an advantage to young,
    genotyped individuals over old animals with only pedigree records
    \cite{Aguilar2009}.
  \item Pedigree depth varies among individuals; some animals have more
    ancestors and therefore more precise estimates \cite{Aguilar2009}.
\end{itemize}

% Problems with SS-BLUP
\begin{itemize}
  \item Weight $w$ for $\mathbf{A}$ and $\mathbf{G}$.
  \item The optimal weight differs between traits \cite{Ashraf2016}.
  \item compatibility between $\mathbf{A}$ and $\mathbf{G}$
    \cite{Christensen2012}
\end{itemize}

% Workings and advantages of the single step procedure.
Independently from each other, \citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a BLUP method that blends the numerator
relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$.

% Physiological epistasis (added benefits of downstream 'omics' predictors)
Compared to pedigree-based prediction, genomic prediction captures Mendelian
sampling, which is the difference between the expected and the realized
relationship between individuals.
Despite its great success and benefits, genomic prediction does not seem to
capture all physiological epistasis \cite{Sackton2016,Reif2015}, which rather 
manifests itself in the form of main and interaction effects of and between 
downstream 'omics' strata, such as the transcriptome, the proteome and the 
metabolome.
This has lead to the exploration of

Besides genomic data, other predictors have proven their value in the
prediction of complex traits in plant populations \cite{Guo2016,Westhues2017}
and in humans \cite{Vazquez2016}.

% Special situation in hybrid breeding.
Hybrid breeding --- with noteworthy commercial applications in pig, chicken and 
maize breeding --- is a particularly challenging application field for
prediction tools.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology \cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.

% -   Mention scarcity of data on single step prediction in plant populations.
Single step prediction has been a hot topic in animal breeding for a few years,
but so far, to the best of our knowledge, has only been considered in plants
only in one study on wheat lines from CIMMYT \cite{Ashraf2016}. 
So far, its applications in hybrid populations are limited to pig breeding
\cite{Xiang2015,Xiang2016,Tusell2016}.
% Are there applications in chicken breeding? (Hints in Christensen et al.
% 2014, GSE).
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to hybrids, (2) explore the utility of single step 
predition with a quantitative predictor and (3) evaluate the impact of the
particular genetic space covered by the incomplete predictor on predictive
ability.

% -   How extensive are pedigree records in commercial breeding companies?

@@

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
% Describe the many hacks that have been developed to tackle problems with BLUP
% models.

@@

\Genetics2level{Superiority of single step models}

\Genetics2level{Coverage of the genetic space}
Animal populations are so large that it does not truly matter, which animals are 
being genotyped, but in plant breeding the situation is different and we would 
be interested in determining which individuals should be complemented with 
information on another predictor.

Prediction of phenotypic performance is particularly crucial in hybrid breeding
programs because the number of potential selection candidates is $n^{2}$ where
$n$ corresponds to the number of parent lines.

Given recent results on the utility of downstream 'omics' predictors in a study
on hybrid prediction in a European hybrid maize panel \cite{Westhues2017}, our 
objectives were i) to explore the benefits of a single step procedure for the 
same material and ii) to examine the influence on the composition of the set of
genotypes that is covered by more than one predictor on the success of single
step prediction.

### 2017-03-27 paper/notes.Rmd
Check whether repo transfer to Github worked
Init manuscript intro

-   Optimized the weight of genomic and pedigree information in blending the
  two matrices by fitting them simultaneous as random effects in a mixed model.

@@

-   "A single-step method extends the usual pedigree-based method by replacing
  the additive relationship matrix constructed from pedigree by an additive
  relatoinship matrix that combines the marker-based relationship matrix for
  genotyped animals with the pedigree-based relationship matrix."

@@

-   "Thus, their method is called single-step BLUP (SS-BLUP), and is expected
  to yield unbiased predictions under multivariate normality, even in
  populations that are undergoing selection and non-random mating. This is
  improtant, because genotypes are usually collected only on superior animals
  and this can lead to biased evaluations. A properly calculated BLUP
  evaluation has been shown to be free of this selection bias [15-21]."

>>>>

-   Optimized the weight of genomic and pedigree information in blending the
  two matrices by fitting them simultaneously as random effects in a mixed 
  model.

@@

-   "A single-step method extends the usual pedigree-based method by replacing
  the additive relationship matrix constructed from pedigree by an additive
  relationship matrix that combines the marker-based relationship matrix for
  genotyped animals with the pedigree-based relationship matrix."

@@

-   "Thus, their method is called single-step BLUP (SS-BLUP), and is expected
  to yield unbiased predictions under multivariate normality, even in
  populations that are undergoing selection and non-random mating. This is
  important, because genotypes are usually collected only on superior animals
  and this can lead to biased evaluations. A properly calculated BLUP
  evaluation has been shown to be free of this selection bias [15-21]."
